Designing Types for R, Empirically

### 2020-09-13 paper/main.tex
lots of prose and comments, can trim. waiting on last data to finish kaggle and revdeps discussion. need to add NA discussion to background

\begin{document}
\title{Designing Types for R, Empirically}
\author{Aviral Goel}\affiliation{\institution{Northeastern University}}
\author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
\author{Jan Vitek}\affiliation{\institution{Norteastern / Czech Technical University}}

@@

 \AT{Kinds of failures: vectors passed to scalars, wrong things passed to classes, nulls passed to non-nullable arguments.}

@@

 First, we found that many of these errors occurred in arguments that have a {\tt NULL} default value.
 This would be the case when the programmer only tested the function by passing values to these null-by-default arguments, and clients of the package make use of the default.
 In fact, we observed \AT{something about coverage.}

 The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or resulting in valid function behaviour that was untested by the original package designer.
 Here, the first case can be explained by a lack of testing (or generally incorrect use in a client), and the second case is explained by programmers not fully understanding R's language semantics (\AT{as who would expect as.integer(NULL) to yield integer(0)?}).

 Next, we were curious about how often vectors were passed to arguments expecting scalars.
 We found \AT{finish analysis, numbers}.
 Besides an outright error, this kind of contract assertion failure might indicate that a function was not well-tested, in that it was only ever tested with unit-length vectors being passed to an argument which is intended to have a vector type.
 Further, these errors may reveal functions that were not designed with a vector-typed argument in mind, but can in fact handle vectors of values (in R, most functions that can accept scalars can also accept vectors).

@@

>>>>

\begin{document}
\title{Designing Types for R, Empirically}
\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
\author{Aviral Goel}\affiliation{\institution{Northeastern University}}
\author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
\author{Jan Vitek}\affiliation{\institution{Norteastern / Czech Technical University}}

@@

@@

 First, we found that many of these errors occurred in arguments that have a {\tt NULL} default value.
 This would be the case when the programmer only tested the function by passing values to these null-by-default arguments, and clients of the package make use of the default.
 As an example, we observed a contract assertion failure when {\tt inner} and {\tt labels} were {\tt NULL} in the following function:
 \begin{lstlisting}
nlme::nfGroupedData <- function (formula, data = NULL, order.groups = TRUE, 
    FUN = function(x) max(x, na.rm = TRUE), outer = NULL, inner = NULL, 
    labels = NULL, units = NULL) {
  # ...
}
 \end{lstlisting}
 In fact, we observed \AT{something about coverage.}

 The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or resulting in valid function behaviour that was untested by the original package designer.
 Here, the first case can be explained by a lack of testing (or generally incorrect use in a client), and the second case is explained by programmers not fully understanding R's language semantics.
 For example, we observed this kind of error in the following function: 
 \begin{lstlisting}
BBmisc::convertIntegers <- function (x) {
    if (is.integer(x)) 
        return(x)
    if (length(x) == 0L || (is.atomic(x) && all(is.na(x)))) 
        return(as.integer(x))
    if (is.numeric(x)) {
        # ...
    }
    return(x)
}
 \end{lstlisting}
Here, if {\tt x} is {\tt NULL}, the second branch of the conditional will be triggered (as in R, \code{length(NULL) == 0}), and the function will return \code{as.integer(NULL)}, which curiously returns \code{integer(0)}, a zero-length vector of integers (one might expect it to return the integer \code{NA} value, or error).

 Next, we were curious about how often vectors were passed to arguments expecting scalars.
 We found \AT{finish analysis, numbers}.
 Besides an outright error, this kind of contract assertion failure might indicate that a function was not well-tested, in that it was only ever tested with unit-length vectors being passed to an argument which is intended to have a vector type.
 Further, these errors may reveal functions that were not designed with a vector-typed argument in mind, but can in fact handle vectors of values (in R, most functions that can accept scalars can also accept vectors).
 As an example, consider the function \code{BBmisc::strrepeat}, which takes a string and repeats it a specified number of times:
 \begin{lstlisting}
BBmisc::strrepeat <- function (x, n, sep = "") {
    paste0(rep.int(x, n), collapse = sep)
}

BBmisc::strrepeat("a", 3) # => "aaa"
BBmisc::strrepeat(c("a", "b"), 3) # => "ababab"
 \end{lstlisting}
 This function was only ever tested with unit-length vectors passed to {\tt x}, even though technically it can handle longer vectors, as per the two sample calls above.
 This could be attributed to poor quality testing, or misunderstanding language semantics (e.g., misunderstanding the semantics of \code{paste0} and \code{rep.int}), but we found other instances of type errors where the functions really ought to have been tested with the offending type, for instance:
 \begin{lstlisting}
combinat::permn <- function (x, fun = NULL, ...) {
    if (is.numeric(x) && length(x) == 1 && x > 0 && trunc(x) == x) 
        x <- seq(x)
    # ...
}
 \end{lstlisting}
 As per the documentation, this function is intended to take a vector, and generate all permutations of elements in that vector.
If given a scalar ``integer'' $n$, it will generate all permutations for the list $[1, 2, ..., n]$.
Curiously, the function was only tested by passing an integer, and not ever with a vector (even though, presumably, that is the main utility of the function).

Finally, we were curious to see what patterns of errors occurred in arguments expecting classes.
\AT{Overall, the numbers are ...}.
Many of these errors appear to be related to ``subclasses'' being passed to arguments expecting a ``superclass''.
For instance, there are \AT{many} instances of \CLASS{{\tt data.frame}} being passed to arguments expecting tibbles or data.tables (which can be readily converted).
\AT{Check dispatch, and maybe count it out when doing this.}

Other common patterns include \CLASS{{\tt matrix}}, \CLASS{{\tt data.frame}}, and vectors being passed to arguments expecting \CLASS{{\tt array}}, which is \AT{elaborate on arrays}.
\AT{Fill rest of this in: vectors and data frames passed to matrices. reiterate how this coercion is ad hoc. maybe talk about coercion semantics.}

@@

\begin{itemize}
  \item compare which functions have failures 
  \item look at the coverage 
  \item types of errors?
\end{itemize}
